---
title: "ESM 244 Lab 1"
author: "Bren Comms Department"
date: "1/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
library(dplyr)
```

### Read in SF trees data

```{r}
sf_trees <- read_csv(here("data", "sf_trees","sf_trees.csv"))
```

### Basic wrangling reminders

Refresh some skills for daata wrangling & summary statistics using functions in the 'dplyr' paackage.

Find the top 5 highest observations of trees by legal status, do some wrangling, and make a graph.
```{r}
# Wrangling on types of tree by legal status
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% #drop NAs in any row where you don't want them based on column variables
  rename(tree_count = n) %>% #new name on the left, old name on the right
  relocate(tree_count) %>% #moves tree_count to the first column position
  #just keep legal_status categories with top tree_count values
  slice_max(tree_count, n = 5) #to identify rows with highest value specified
  
```

Make a graph of those top 5 obsesrvations by legal status.
```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + #fct_reorder can be reversed (asc vs. desc)
  geom_col() +
  labs(x = "Legal Status",
       y = "Tree Count") +
  coord_flip() + #flips the x and y axes
  theme_minimal()
```

### A few more data wraangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia.

```{r}
blackwood_acacia_Psite <- sf_trees %>% 
  filter(legal_status == "Permitted Site") #fxn we'll use the most in wrangling data

# Exaample: keep any observations where a certain string is detected anywhere within that variable for an observation
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, latitude, longitude) #useful for picking/excluding columns, helpful when there are too many

#Plot trees! (A preview! Because R doesn't know these are geographic points yet)
ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) +
  geom_point()
```

### tidyr::separate() and unite() functions

Useful for combining or separating fxns.

```{r}
#split columns by separators
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific","spp_common"), sep = "::")
  # separate(column, into = c("new column name 1", "new column name 2"), sep = "type of separator")
  
```

Example: tidyr::unite()

Combine tree and legal_status columns
```{r}
sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep = "_cool!_")
  # unite("name of new united column", column 1 to unite:column 2 to unite, sep = "type of separator")
```


